Put Title here

### 2018-12-11 General WLD combinatorics.tex
Minor notes on section 2.2 and 2.3

We first give a quick summary of the matroid terminology that we will need; it is not intended as a comprehensive introduction to matroids and the interested reader is referred to [ref].

@@

In particular, this allows us to prove results about $M(W)$ by considering the behavior of propagators in $W$. We record a few elementary facts about the rank and cycles of $M(W)$ here as an example of this.\note{It would be nice to have a few more; do we need any others, and/or are there any cool ones?}

\begin{lem}\label{lem facts about WLD matroids}
Let $W = (\cP,n)$ be an admissible Wilson loop diagram. Then:
\begin{enumerate}
\item The rank of a set $V \subseteq [n]$ is bounded above by $\min\{|V|,|\Prop(V)|\}$, with $\rk(V) = |V|$ if and only if $V$ is an independent set.
%\item Let $v \in [n]$ and $q \in \Prop(v)$. Then for any $V \subset [n]$ such that $q \not \in \Prop(V)$, we have $\rk(V \cup \{v\}) = \rk(V) +1$, i.e. adding a vertex that supports a new propagator to a set increases the rank of the set.
\item If $C \subseteq [n]$ is a cycle, then $\rk(C) = |Prop(C)|$.
\item If $[n]$ can be partitioned into at least two non-empty sets, each of which support different sets of propagators, \bas [n] = \cup_i V(P_i) \quad \textrm{s.t.} \quad \cup P_i = \cP; \quad  P_i \cap P_j = \emptyset \;, \eas the the matroid $M(W)$ is seperable, \bas M(W) = \bigoplus_i M(P_i, V(P_i)) \;.\eas 
\end{enumerate}
\end{lem}
\begin{proof}
The first part of (1) is \cite[Equation (9)]{wilsonloop} and surrounding discussion, and the second part is standard matroid theory. (2) is \cite[Lemma 3.27]{wilsonloop}. (3) is also in \cite[Lemma FIND]{wilsonloop}.
\end{proof}

@@

\begin{proof}
We begin by giving an algorithm for the decomposition, then prove its uniqueness. Let $W = (\cP, n)$, with $|\cP| = k$.

@@

Since $W$ is admissible, by Lemma \ref{tausimpleplanarlem} $\tau(W)$ is a simple graph. Furthermore, $t$ is not all of $\tau(W)$.

Let $t$ be a triangulated piece of the decomposition of $\tau(W)$ given in Lemma \ref{decompositionlem}; note that $t$ cannot be equal to $\tau(W)$ by the definition of admissible diagrams.

If $t$ has 2 vertices, then $t$ has two connected components. Therefore, $t$ corresponds to a propagator that connects two non-adjacent edges. Therefore, the trivial triangulation is a trivial exact subdiagram.

Now suppose that $t$ has $m>2$ vertices; we count how many edges of $t$ are not on the outer face of $\tau(W)$.  Consider the intersection of $t$ with the outer face of $\tau(W)$: this is a possibly disconnected subgraph of the polygon of $\tau(W)$ with $m$ vertices, call it $S$.  Let $k$ be the number of connected components of $S$.   To join the components $S$ must have $k$ edges not in the outer face of $\tau(W)$.  Furthermore $t$ has $m-3$ edges not in its outer face and so also not in the outer face of $\tau(W)$.  Thus there are $m-3+k$ edges of $t$ not in the outer face of $\tau(W)$.

Each of these $m-3+k$ internal edges correspond to propagators in $W$, call this set $P$.  Next we count the size of $V(P)$.  Each of the $m$ vertices in the outer face of $t$ corresponds to an edge of $W$. These $m$ edges define $k$ connected compnents of the outer polygon of $W$. Thus the set $V(P)$ has $m+k$ vertices.  In other words, 
\[|V(P)| = m+k = |P| -3\;.\]

@@

Suppose the set $V(P)$ defines $c$ disjoint cyclic intervals of $[n]$. Then the corresponding subgraph of $\tau(W)$ has $m-c$ vertices. Call it $t$. If $t$ were a triangulated component, $t$ would have $k -c$ internal edges.

The graph $t$ has $k$ edges that come from propagators, and $m - 2c$ edges that come from the boundary polygon of $\tau(W)$. Since $t$ has $m-c$ vertices, it has $m-c$ external edges, of which $c$ come from propagators. Therefore, of the $k$ edges of $t$ that come from propagators, $k-c$ are internal to the connected component. Therefore, this is triangulated.

@@

\begin{cor} \label{uniqueproppartitioncor}
Any admissible Wilson loop diagram $W = (\cP, n)$ can be uniquely decomposed into maximal exact subdiagrams. These maximal subdiagrams partition $\cP$.
\end{cor}

@@

\begin{lem} \label{contractsubdiaglem} \cite[Theorem 3.33]{wilsonloop} 
Let $W = (\cP, n)$ be an admissible Wilson loop diagram and $P \subseteq \cP$. If the set $V(P)^c$ has rank $|P^c|$, then the matroid defined by the subdiagram $(P, V(P))$ is equal to the contraction $M(W)/V(P)^c$.
\end{lem}

@@

\begin{dfn}\label{def prop flat}
Let $W = (\cP,n)$ be an admissible Wilson loop diagram and $P \subseteq \cP$. The set $F(P) := V(P^c)^c$ is called the {\em propagator flat} of $P$.
\end{dfn}
Thus in Lemma \ref{contractsubdiaglem} above we would contract $M(W)$ by the propagator flat $F(P^c)$ of the {\em complement} of $P$ in order to study $(P,V(P))$. The justification for this notation is given by (1) of the next lemma: the set $F(P)$ consists of exactly those vertices which {\em only} support propagators in $P$.

\begin{lem}\label{lem properties of prop flats}
Let $F(P)$ be a propagator flat as defined above. Then
\begin{enumerate}
\item $F(P) = V(P) \setminus V(P^c)$.
\item If $Q\subseteq P$ then $F(Q) \subseteq F(P)$.
\item $F(P)$ is a flat of $M(W)$, thus justifying the name ``propagator flat''.
\end{enumerate}
\end{lem}
\begin{proof}
The proof of (1) and (2) are routine applications of the definition and are omitted.

@@

Let $v \in C$. Clearly $\Prop(v) \subseteq \Prop(C)$, and so by Lemma \ref{lem properties of prop flatfs} we have $F(\Prop(v)) \subseteq F(\Prop(C))$. Since $v \in F(\Prop(v))$ by the definition of propagator flat, we have $v \in F(\Prop(C))$ as required.

@@

For any $e \not\in F$, we certainly have $\rk((F\setminus C)\cup\{e\}) = \rk(F\setminus C) +1$ since $F$ is a flat. Now let $e \in C$, and suppose that $\rk((F\setminus C)\cup\{e\}) = \rk(F\setminus C)$. This implies that $(F\setminus C)\cup \{e\}$ is dependent, and hence contains a circuit. This circuit must contain at least two elements (since all vertices support at least one propagator), but this contradicts the fact that $C$ was the union of all circuits in $F$.

@@

\begin{lem} \label{maxexactcomplementrank}
Let $W = (\cP, n)$ be a Wilson loop diagram, and $P \subseteq \cP$. Then: \begin{enumerate}
\item If $(P,V(P))$ is an exact subdiagram in $W$, then $\rk(F(P^c)) = |P^c|$.
\item If $(P,V(P))$ is a maximal exact subdiagram in $W$, then $F(P^c)$ is a cyclic flat.
\end{enumerate}
\end{lem}

Note that by definition, since $F(P^c)$ has at $|P^c|$ propagators in it, this is an upper bound on the rank. However, showing equality on the first part is harder.

\begin{proof}
We first verify that $F(P^c) = V(P)^c$ is non-empty and dependent for any exact subdiagram $(P,V(P))$. The definition of exactness means that the vertices $V(P)$ cannot support a larger set of propagators (otherwise $W$ would no longer be admissible). Therefore, since $P \subsetneq \cP$, there must be at least one propagator supported on $V(P)^c$, i.e. it cannot be empty.

Since $W$ is admissible, we have $n \geq |\cP| + 4$. Rewriting this as
\[|V(P)| + |F(P^c)|  \geq  |P| + |P^c| +4,\]
and combining it with the fact that $|V(P)| = |P| + 3$ (from the exactness of $(P,V(P))$), we obtain
\begin{equation}\label{eq F is dependent}|F(P^c)| > |P^c| \;.\end{equation}
By Lemma \ref{lem properties of prop flats}, $F(P^c)$ is the set of vertices that {\em only} supports propagators in $P^c$, so in particular we have $\Prop(F(P^c)) = P^c$. Equation \eqref{eq F is dependent} is therefore saying that $F(P^c)$ supports fewer propagators than the number of vertices it contains, i.e. $F(P^c)$ is dependent.

We address part (2) first. Suppose that $(P, V(P))$ is a maximal exact subdiagram; by Corollary \ref{classifyflats} we can decompose $F(P^c)$ as
\begin{equation}\label{eq decompose F}F(P^c) = C \sqcup S,\end{equation}
where $C$ is the largest cyclic flat contained in $F(P^c)$ and $S$ is an  independent set. Note that $C$ must be non-empty since $F(P^c)$ is dependent. We proceed by showing that $S$ must be empty, forcing $F(P^c)$ to be a cyclic flat.

Since $C = F(\Prop(C))$ by Lemma \ref{lem decompose flat}, define $Q := \Prop(C)^c$ so that we may write $C = F(Q^c) = V(Q)^c$. Note that since $C \subseteq F(P^c)$, $P \subseteq Q$. Furthermore, we have 
\begin{equation}\rk(S) = |S| \leq \rk(F(P^c)) - \rk(C)  \leq |P^c| - \rk(C) \;, \label{increasingrank}\end{equation}
where first equality comes from the fact that $S$ is an independent set, and the final inequality comes from Lemma \ref{lem facts about WLD matroids}(1). Since $\rk(C) = |\Prop(C)| = |Q^c|$ by Lemma \ref{lem facts about WLD matroids}(2), we can rearrange \eqref{increasingrank} to obtain the inequality
\begin{equation} |P| + |S| \leq |P| + |P^c| - \rk(C) = |Q| \;. \label{Qsizebound}\end{equation}
Furthermore, since $V(Q) = S \sqcup V(P)$ (by equation [ref] and the definition of $Q$) we may write  
\begin{equation} |V(Q)| = |S| + |V(P)| = |S| + |P| +3 \label{VQsize} \;.\end{equation} 
Combining this with \eqref{Qsizebound} gives $|V(Q)| - 3 \leq |Q|$. That is, either $W$ was not admissible, or $(P,V(P))$ was not a maximal exact subdiagram. In other words, $S$ cannot be non-empty, proving that $F(P^c)$ is a cyclic flat, and therefore of maximal rank.

Let $P$ be a maximal exact subdiagram as above, and let $(R, V(R))$ be an exact subdiagram that is not maximal. That is, $R \subsetneq P$.

@@

\begin{thm} \label{exactcircuitlem}
Let $W':= (P, V(P))$ be a subdiagram of an admissible Wilson loop diagram $W= (\cP, n)$. Then $W'$ is an exact subdiagram if and only if $M(W')$ is a uniform matroid of rank $|P|$.
\end{thm}

>>>>

We first give a quick summary of the matroid terminology that we will need; it is not intended as a comprehensive introduction to matroids and the interested reader is referred to [ref][find a good matroid reference].

@@

In particular, this allows us to prove results about $M(W)$ by considering the behavior of propagators in $W$. We record a few elementary facts about the rank and cycles of $M(W)$ here as an example of this.

\begin{lem}\label{lem facts about WLD matroids}
Let $W = (\cP,n)$ be an admissible Wilson loop diagram. Then:
\begin{enumerate}
\item The rank of a set $V \subseteq [n]$ is bounded above by $\min\{|V|,|\Prop(V)|\}$, with $\rk(V) = |V|$ if and only if $V$ is an independent set.
%\item Let $v \in [n]$ and $q \in \Prop(v)$. Then for any $V \subset [n]$ such that $q \not \in \Prop(V)$, we have $\rk(V \cup \{v\}) = \rk(V) +1$, i.e. adding a vertex that supports a new propagator to a set increases the rank of the set.
\item If $C \subseteq [n]$ is a cycle, then $\rk(C) = |Prop(C)|$.
\item If $[n]$ can be partitioned into at least two non-empty sets, each of which support different sets of propagators, \bas [n] = \cup_i V(P_i) \quad \textrm{s.t.} \quad \cup P_i = \cP; \quad  P_i \cap P_j = \emptyset \;, \eas the the matroid $M(W)$ is seperable, \bas M(W) = \bigoplus_i M(P_i, V(P_i)) \;.\eas \note{Closest reference I could find for this was Theorem 3.23 of Agarwala/Amit, which relates to $F(P_i)$ instead of $V(P_i)$; check if there's a better reference?}
\end{enumerate}
\end{lem}
\begin{proof}
The first part of (1) is \cite[Equation (9)]{wilsonloop} and surrounding discussion, and the second part is standard matroid theory. (2) is \cite[Lemma 3.27]{wilsonloop}. (3) is also in \cite[Lemma FIND]{wilsonloop}.
\end{proof}

@@

\begin{proof}
We begin by giving an algorithm for the decomposition, then prove its uniqueness. Let $W = (\cP, [n])$, with $|\cP| = k$.

@@

Since $W$ is admissible, by Lemma \ref{tausimpleplanarlem} $\tau(W)$ is a simple graph. Let $t$ be a triangulated piece of the decomposition of $\tau(W)$ given in Lemma \ref{decompositionlem}; note that $t$ cannot be equal to $\tau(W)$ by the definition of admissible diagrams.

If $t$ has 2 vertices, \hlfix{then $t$ has two connected components.}{I'm not sure what this is meant to say?} Therefore, $t$ corresponds to a propagator that connects two non-adjacent edges. Therefore, the trivial triangulation is a trivial exact subdiagram.

Now suppose that $t$ has $m>2$ vertices; we count how many edges of $t$ are not on the outer face of $\tau(W)$.  Consider the intersection of $t$ with the outer face of $\tau(W)$: this is a possibly disconnected subgraph of the polygon of $\tau(W)$ with $m$ vertices. Call this new subgraph $S$, and let $j$ be the number of connected components of $S$.   To join the components \hlfix{$S$}{check: is this $S$ or $t$?} must have $j$ edges not in the outer face of $\tau(W)$.  Furthermore $t$ has $m-3$ edges not in its outer face and so also not in the outer face of $\tau(W)$.  Thus there are $m-3+j$ edges of $t$ not in the outer face of $\tau(W)$.

Each of these $m-3+j$ internal edges correspond to a propagator in $W$; call this set of propagators $P$.  Next we count the size of $V(P)$.  Each of the $m$ vertices in the outer face of $t$ corresponds to an edge of $W$. These $m$ edges define $j$ connected components of the outer polygon of $W$. Thus the set $V(P)$ has $m+j$ vertices.  In other words, 
\[|V(P)| = m+j = |P| +3\;.\]
Thus the subdiagram $(P,V(P))$ defined by $t$ is exact.

@@

Suppose the set $V(P)$ defines $c$ disjoint cyclic intervals of $[n]$. Then the corresponding subgraph of $\tau(W)$ has $m-c$ vertices. \hlfix{Call it $t$.}{$t$ is defined twice, and I'm not sure what $m$ and $j$ are for this direction of the proof.} If $t$ were a triangulated component, $t$ would have $j -c$ internal edges.

The graph $t$ has $j$ edges that come from propagators, and $m - 2c$ edges that come from the boundary polygon of $\tau(W)$. Since $t$ has $m-c$ vertices, it has $m-c$ external edges, of which $c$ come from propagators. Therefore, of the $j$ edges of $t$ that come from propagators, $j-c$ are internal to the connected component. Therefore, this is triangulated.

@@

\begin{cor} \label{uniqueproppartitioncor}
Any admissible Wilson loop diagram $W = (\cP, [n])$ can be uniquely decomposed into maximal exact subdiagrams. These maximal subdiagrams partition $\cP$.
\end{cor}

@@

\begin{lem} \label{contractsubdiaglem} \cite[Theorem 3.33]{wilsonloop} 
Let $W = (\cP, [n])$ be an admissible Wilson loop diagram and $P \subseteq \cP$. If the set $V(P)^c$ has rank $|P^c|$, then the matroid defined by the subdiagram $(P, V(P))$ is equal to the contraction $M(W)/V(P)^c$.
\end{lem}

@@

\begin{dfn}\label{def prop flat}
Let $W = (\cP,[n])$ be an admissible Wilson loop diagram and $P \subseteq \cP$. The set $F(P) := V(P^c)^c$ is called the {\em propagator flat} of $P$.
\end{dfn}
Thus in Lemma \ref{contractsubdiaglem} above we would contract $M(W)$ by the propagator flat $F(P^c)$ of the {\em complement} of $P$ in order to study $(P,V(P))$. The justification for this notation is given by (1) of the next lemma: the set $F(P)$ consists of vertices which {\em only} support propagators in $P$ (or no propagators at all).

\begin{lem}\label{lem properties of prop flats}
Let $F(P)$ be a propagator flat as defined above. Then
\begin{enumerate}
\item $F(\emptyset)$ is exactly the set of vertices which support no propagators, while for $P \neq \emptyset$ we have \[F(P) = \big(V(P) \setminus V(P^c)\big) \cup F(\emptyset).\] \note{updated (1) to match new definition of $F(P)$; }
\item If $Q\subseteq P$ then $F(Q) \subseteq F(P)$.
\item $F(P)$ is a flat of $M(W)$, thus justifying the name ``propagator flat''.
\end{enumerate}
\end{lem}
\begin{proof}
The proof of (1) and (2) are routine applications of the definition and are omitted.

@@

Let $v \in C$. Clearly $\Prop(v) \subseteq \Prop(C)$, and so by Lemma \ref{lem properties of prop flats} we have $F(\Prop(v)) \subseteq F(\Prop(C))$. Since $v \in F(\Prop(v))$ by the definition of propagator flat, we have $v \in F(\Prop(C))$ as required.

@@

For any $e \not\in F$, we certainly have $\rk((F\setminus C)\cup\{e\}) = \rk(F\setminus C) +1$ since $F$ is a flat. Now let $e \in C$, and suppose that $\rk((F\setminus C)\cup\{e\}) = \rk(F\setminus C)$. This implies that $(F\setminus C)\cup \{e\}$ is dependent, and hence contains a circuit. This circuit must contain at least two elements \note{What if $e \in C$ supported no propagators? $\{e\}$ would then be a circuit on its own}(since all vertices support at least one propagator), but this contradicts the fact that $C$ was the union of all circuits in $F$.

@@

\begin{lem} \label{maxexactcomplementrank}
Let $W = (\cP, [n])$ be a Wilson loop diagram, and $P \subseteq \cP$. Then: \begin{enumerate}
\item If $(P,V(P))$ is an exact subdiagram in $W$, then $\rk(F(P^c)) = |P^c|$.
\item If $(P,V(P))$ is a maximal exact subdiagram in $W$, then $F(P^c)$ is a cyclic flat.
\end{enumerate}
\end{lem}

Note that $|P^c|$ is always an upper bound on the rank of $F(P^c)$ for any $P \subseteq \cP$, since $F(P^c)$ supports at most $P^c$ propagators. However, showing equality is far harder.

\begin{proof}
First note that if $(P,V(P))$ is exact then the admissiblity of $W$ guarantees that $F(P^c)$ is non-empty.

Since $W$ is admissible, we have $n \geq |\cP| + 4$. Rewriting this as
\[|V(P)| + |F(P^c)|  \geq  |P| + |P^c| +4,\]
and combining it with the fact that $|V(P)| = |P| + 3$ (from the exactness of $(P,V(P))$), we obtain
\begin{equation}\label{eq F is dependent}|F(P^c)| > |P^c| \;.\end{equation}
By Lemma \ref{lem properties of prop flats}, $F(P^c)$ is a set of vertices that supports only propagators in $P^c$, so in particular we have $\Prop(F(P^c)) = P^c$. Equation \eqref{eq F is dependent} is therefore saying that $F(P^c)$ supports fewer propagators than the number of vertices it contains, i.e. $F(P^c)$ is dependent.

We address part (2) first. Suppose that $(P, V(P))$ is a maximal exact subdiagram; by Corollary \ref{classifyflats} we can decompose $F(P^c)$ as
\begin{equation}\label{eq decompose F}F(P^c) = C \sqcup S,\end{equation}
where $C$ is the largest cyclic flat contained in $F(P^c)$ and $S$ is an  independent set. Note that $C$ must be non-empty since $F(P^c)$ is non-empty and dependent. We proceed by showing that $S$ must be empty, forcing $F(P^c)$ to be a cyclic flat.

Since $C = F(\Prop(C))$ by Lemma \ref{lem decompose flat}, define $Q := \Prop(C)^c$ so that we may write $C = F(Q^c) = V(Q)^c$. Note that since $C \subseteq F(P^c)$, $P \subseteq Q$. Furthermore, we have 
\begin{equation}|S| = \rk(S) \leq \rk(F(P^c)) - \rk(C)  \leq |P^c| - \rk(C) \;, \label{increasingrank}\end{equation}
where first equality comes from the fact that $S$ is an independent set,\note{Did we ever address the question of where the middle inequality comes from? Can we argue by number of propagators?} and the final inequality comes from Lemma \ref{lem facts about WLD matroids}(1). Since $\rk(C) = |\Prop(C)| = |Q^c|$ by Lemma \ref{lem facts about WLD matroids}(2), we can rearrange \eqref{increasingrank} to obtain the inequality
\begin{equation} |P| + |S| \leq |P| + |P^c| - \rk(C) = |Q| \;. \label{Qsizebound}\end{equation}
Furthermore, since $V(Q) = S \sqcup V(P)$ (by equation \eqref{eq decompose F} and the definition of $Q$) we may write  
\begin{equation} |V(Q)| = |S| + |V(P)| = |S| + |P| +3 \label{VQsize} \;.\end{equation} 
Combining this with \eqref{Qsizebound} gives $|V(Q)| \leq |Q| + 3$. Since $W$ is admissible, we conclude that $(Q,V(Q))$ is exact; since $V(Q) = V(P) \sqcup S$ and $(P,V(P))$ is maximal exact, it follows that $S = \emptyset$. It now follows from \eqref{eq decompose F} that $F(P^c)$ is a cyclic flat, and hence from Lemma \ref{lem facts about WLD matroids} that
\[\rk(F(P^c)) = |Prop(F(P^c))| = |P^c|.\]

Let $P$ be a maximal exact subdiagram as above, and let $(R,V(R))$ be an exact subdiagram that is not maximal. That is, $R \subsetneq P$.

@@

\begin{thm} \label{exactcircuitlem}
Let $W':= (P, V(P))$ be a subdiagram of an admissible Wilson loop diagram $W= (\cP, [n])$. Then $W'$ is an exact subdiagram if and only if $M(W')$ is a uniform matroid of rank $|P|$.
\end{thm}
